---
title: "An analysis of my own chess playing habits"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(tidyverse)
library(hrbrthemes)
library(readxl)
library(DT)
library(pander)
library(ggpubr)


result <- read_xlsx("chess_games_2020-08-21_to_2022-05.29.xlsx")
```

This project focuses on basic analysis (playing habits, win-percentages, etc) of my chess games from 21/08/2020 to 29/05/2022. In this period I have played a total of 2539 games on Chess.com, which is where I have gotten my games data-set from using the R package "bigchess". 

## Basic Stats

<center>
```{r basic stats, fig.dim = c(7, 5)}

#games as white --------
games_as_white <- result %>% filter(White == "Shamborambo") %>%
  count() %>% as.integer()

#games as black ---------
games_as_black <- result %>% filter(Black == "Shamborambo") %>%
  count() %>% as.integer()

#outcomes as white ---------
wins_as_white <- result %>%
  filter(White == "Shamborambo" & Result == "1-0") %>% 
  count() %>% as.integer()

losses_as_white <- result %>%
  filter(White == "Shamborambo" & Result == "0-1") %>% 
  count() %>% as.integer()

draws_as_white <- result %>%
  filter(White == "Shamborambo" & Result == "1/2-1/2") %>% 
  count() %>% as.integer()

#outcomes as black ---------
wins_as_black <- result %>%
  filter(Black == "Shamborambo" & Result == "0-1") %>% 
  count() %>% as.integer()

losses_as_black <- result %>%
  filter(Black == "Shamborambo" & Result == "1-0") %>% 
  count() %>% as.integer()

draws_as_black <- result %>%
  filter(Black == "Shamborambo" & Result == "1/2-1/2") %>% 
  count() %>% as.integer()

#win/draw/loss rates as white ---------
win_rate_as_white <- round((wins_as_white/games_as_white)*100, 2)
draw_rate_as_white <- round((draws_as_white/games_as_white)*100, 2)
loss_rate_as_white <- round((losses_as_white/games_as_white)*100,2)

#win/draw/loss rates as black ---------
win_rate_as_black <- round((wins_as_black/games_as_black)*100, 2)
draw_rate_as_black <- round((draws_as_black/games_as_black)*100, 2)
loss_rate_as_black <- round((losses_as_black/games_as_black)*100,2)

#totals -------
wins_total <- result %>% 
  filter(White == "Shamborambo" & Result == "1-0" | Black == "Shamborambo" & Result == "0-1") %>% 
  count() %>% as.integer()

win_rate_total <- round((wins_total/2539)*100, 2)

losses_total <- result %>% 
  filter(White == "Shamborambo" & Result == "0-1" | Black == "Shamborambo" & Result == "1-0") %>% 
  count() %>% as.integer()

loss_rate_total <- round((losses_total/2539)*100, 2)

draws_total <- result %>% 
  filter(Result == "1/2-1/2") %>% 
  count() %>% as.integer()

draw_rate_total <- round((draws_total/2539)*100, 2)



#create basic_stats table 
basic_stats <- data.frame(Colour = c("White", "Black", "Total"),
                          Games = c(games_as_white, games_as_black, 2539),
                          Wins = c(wins_as_white, wins_as_black, wins_total),
                          "Win Rate" = c(win_rate_as_white, win_rate_as_black, win_rate_total),
                          Losses = c(losses_as_white, losses_as_black, losses_total),
                          "Loss Rate" = c(loss_rate_as_white, loss_rate_as_black, loss_rate_total),
                          Draws = c(draws_as_white, draws_as_black, draws_total), 
                          "Draw Rate" = c(draw_rate_as_white, draw_rate_as_black, draw_rate_total))
pander(basic_stats)

```
</center>

```{r, fig.align='center'}
#for notmal win rate bar graph
# basic_stats %>% ggplot(aes(x = Colour, y = Win.Rate, fill = Colour)) + 
#   geom_bar(position="dodge", stat="identity", width = 0.5) + coord_flip() + theme_minimal() +
#   scale_fill_brewer(palette = "Set2", guide = "none") +
#   xlab("") + ylab("") + ggtitle("Win Rates") + 
#   theme(plot.title = element_text(hjust = 0.5))

#rework basic stats for grouped bar graph
t <- data.frame(colour = c("White", "Black", "White", "Black", "White", "Black"),
                result = c("Win", "Win", "Draw", "Draw", "Loss", "Loss"),
                rate = c(win_rate_as_white, win_rate_as_black, draw_rate_as_white, draw_rate_as_black, loss_rate_as_white, loss_rate_as_black))
t %>% ggplot(aes(x = rate, y = colour, fill = result)) + 
  geom_bar(position="dodge", stat="identity", width = 0.5) + theme_minimal() +
  xlab("") + ylab("") + ggtitle("Outcome Rates") + 
  theme(plot.title = element_text(hjust = 0.5)) + scale_fill_brewer(palette = "Set2")

```

It's easy to see I have a higher win-rate as white than do I as black but the draw-rate difference between the two is almost negligible. As white, I have drawn only 5.03% of games while as black, I've drawn 5.13% of games. 

I actually tend to lose more as black than I win, which is not great and calls for increased attention to my black openings. 

### Playing Habits

As somebody who plays *almost* daily, it's interesting for me to observe my playing habits. 

```{r games played per day, fig.dim = c(10, 5)}
#time series visualization for games played per day

#games per day
games_per_day <- result %>% group_by(Date) %>% count()
games_per_day$Date <- format(as.Date(games_per_day$Date, '%Y.%m.%d'), "%Y/%m/%d")
games_per_day$Date <- as.Date(games_per_day$Date)

games_per_day %>%
  ggplot(aes(x=Date, y=n)) +
  geom_line(group = 1, color= "steelblue") +
  scale_x_date(date_labels = "%m-%Y", date_breaks = "2 month") +
  theme_pubclean() + 
  theme(axis.text.x=element_text(angle=60, hjust=1),
        plot.title = element_text(hjust = 0.5)) +
  xlab("") + ylab("")

```

### Do I play more on some days than others? 

I was curious to see if I had played more games on some days than I had on others but most importantly, if there was a noticeable difference between my most and least preferred day. 

<center>

```{r, }
#what days do I play the most on? --------

#change Date col to y/m/d format
result$Date <- format(as.Date(result$Date, '%Y.%m.%d'), "%Y/%m/%d")

#games per day
games_per_day$Day <- wday(games_per_day$Date, label=TRUE, abbr = FALSE)

days_and_games_count <- games_per_day %>%
  group_by(Day) %>%
  summarise(n = sum(n)) %>%
  arrange(desc(n)) %>% 
  rename(Games = n)

pander(days_and_games_count)
```

</center>


Prior to this, I had assumed that I had played the most on the weekend, Saturdays and Sundays, but the data tells me that that assumption was wrong. Historically, I've played most of my games on Tuesdays (441 games) and the least (!) on Saturdays (214). I've played 227 more games on a work day than I have on Saturdays ... I'll leave that up to interpretation.

```{r, fig.align='center', fig.dim=c(7,4)}
#visualization 
days_and_games_count %>% ggplot(aes(x = Day, y = Games, fill = Day)) +
  geom_col() + 
  scale_fill_brewer(palette = "Accent", guide = "none") +
  theme_minimal() +
  xlab("") + ylab("")

```

### Do I win more on some days as compared to others?

Another question on my mind when starting this project was to see if I tended to win/lose more on certain days and to try and link performances to my activities for that day. On Saturdays, I play football in the morning and walk back almost 2 hours home post-football. Does that have an impact on my chess games when I *do* play later on? Of course, this isn't a foolproof way to understand the relation but it does provide a simple launching pad for further thought.

<center>
```{r, echo=FALSE, include=FALSE}

#wins
wins <- result %>% 
  filter(White == "Shamborambo" & Result == "1-0" | Black == "Shamborambo" & Result == "0-1") %>% 
  select(Date, White, Black, Result) %>% 
  mutate(Outcome = "win") %>%
  group_by(Date) %>%
  count()

wins$Day <- wday(wins$Date, label=TRUE, abbr = FALSE)

wins_days <- wins %>%
  group_by(Day) %>%
  summarise(n = sum(n)) %>%
  arrange(desc(n)) %>% 
  rename(Wins = n)

#losses
losses <- result %>% 
  filter(White == "Shamborambo" & Result == "0-1" | Black == "Shamborambo" & Result == "1-0") %>% 
  select(Date, White, Black, Result) %>% 
  mutate(Outcome = "loss") %>%
  group_by(Date) %>%
  count()

losses$Day <- wday(losses$Date, label=TRUE, abbr = FALSE)

losses_days <- losses %>%
  group_by(Day) %>%
  summarise(n = sum(n)) %>%
  arrange(desc(n)) %>% 
  rename(Losses = n) 


#draws
draws <- result %>% 
  filter(Result == "1/2-1/2") %>% 
  select(Date, White, Black, Result) %>% 
  mutate(Outcome = "Draw") %>%
  group_by(Date) %>%
  count()

draws$Day <- wday(draws$Date, label=TRUE, abbr = FALSE)

draws_days <- draws %>%
  group_by(Day) %>%
  summarise(n = sum(n)) %>%
  arrange(desc(n)) %>% 
  rename(Draws = n) 


outcomes_days <- left_join(wins_days, losses_days) 
outcomes_days <- left_join(outcomes_days, draws_days)
```

```{r}
chances_days <- data.frame(Day = outcomes_days$Day,
                                      Win.Chances = c(50.57, 52.74, 48.02, 48.52, 48.34, 46.88, 48.6),
                                      Loss.Chances = c(44.9,42.09,47.55,44.33,47.31,48.44,46.26),
                                      Draw.Chances = c(4.54, 5.22,4.43,7.14,4.35,4.69,5.14))

pander(chances_days)
```
</center>

Perhaps I'm not as tired on Saturday post-football as I am on Sunday when my body is sore from the day before? As evident, I win the most on Mondays and the least on Sundays. On the other side of that discussion, I'm least likely to lose on a Monday and most likely to lose on a Sunday, perhapys lending to my tired theory? That's now a question for another project.  


```{r, fig.align='center', fig.dim=c(7,5)}

#visualization 
chances_days %>% ggplot() +
  geom_line(aes(x = Day, y = Win.Chances), group = 1, color = "steelblue") + 
  geom_point(aes(x = Day, y = Win.Chances), color = "steelblue", size = 2.5) + 
  geom_line(aes(x = Day, y = Loss.Chances), group = 1, color = "red") +
  geom_point(aes(x = Day, y = Loss.Chances), color = "red", size = 2.5) + 
  theme_minimal() +
  ylim(c(40,60)) +
  ylab("") + xlab("") + ggtitle("Chances of winning or losing on any given day") +
  theme(plot.title = element_text(hjust = 0.5))

```
In fact, I'm more likely to lose on a Sunday than I am to win - should I stop playing on Sundays? 

